---
title: "Analysis of UK travel to school data"
author: "Robin Lovelace"
date: "1 October 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("setup.R")
```

```{r, eval=FALSE}
source("R/load-all-data.R")
```

# School level data

There are 21,523 schools reported in the 2011 schools census.
Of those, 3,398 met the selection criteria for this analysis of having at least 100 full time 11 to 18 year old pupils.
3.1 million students in this age range attended these schools.
The top 5 of these are presented below

```{r, eval=FALSE}
countvars_ft = grep(pattern = "FT", names(sld11))
sld11$ft = rowSums(sld11[,countvars_ft])
# hist(sld11$ft)
vname_secondary_f = paste0("LEA11_FT_Girls_", 11:18)
vname_secondary_m = paste0("LEA11_FT_Boys_", 11:18)
vname_secondary = c(vname_secondary_f, vname_secondary_m)
sld11$Secondary = rowSums(sld11[,vname_secondary])
sld = sld11[sld11$Secondary >= 100,]
nrow(sld)
sum(sld$Secondary)
sld11_100 = top_n(x = sld11, n = 100, wt = sld11$Secondary)  %>%
  arrange(desc(Secondary))
dir.create("private_data")
saveRDS(sld11_100, "private_data/sld11_100.Rds")
```


```{r}
sld11_100 = readRDS("private_data/sld11_100.Rds")
knitr::kable(
  sld11_100 %>%
    select(LEA11_SchoolName, LEA11_Form7_School_Type_Desc, Total = Secondary) %>%
    slice(1:5)
)
```

# Preliminary analysis of top 100 schools

Preliminary analysis was done on the top 100 schools by number of pupils to test the method without loading all the data.

```{r}

```


spatial distribution is illustrated below.

# OD level data